---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.path = "tools/readme/",
  dpi=300
)
devtools::load_all()
```

# abdiv

This package re-implements measures of ecological diversity from several other
software packages, including `vegan`, `scikit-bio`, and `GUniFrac`.

## Installation

You can install abdiv from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("kylebittinger/abdiv")
```

## Example

Let's say we've surveyed a field and counted the number of plants in each of
two sites.  We've found five species in total, and we'd like to summarize
the diversity of the two sampling sites.  Here are the results for site 1 and
site 2, represented as vectors.

```{r}
s1 <- c(2, 5, 16, 0, 1)
s2 <- c(0, 0, 8, 8, 8)
```

The two sites have about the same number of total plants, but the ditsribution
of species is much different.  More than half the plants in site 1 belong to a
single species, whereas the plants in site 2 are almost evenly distributed
across three different species.  To get started, let's look at a few measures
of diversity for each sample.

Richness measures the total number of species in each sample.

```{r}
richness(s1)
richness(s2)
```

The Shannon index measures both the number of species and the evenness of the
relative abundance values. Site 2 has fewer species, but each species has the
same relative abundance, so the Shannon index is higher.

```{r}
shannon(s1)
shannon(s2)
```

Let's summarize the diversity of site 1 and site 2 using all the functions
available in this library.  The full set of within-sample diversity functions
is available as a character vector in `alpha_diversities`. The term
"α-diversity" means the diversity within a single sample.

```{r alpha-diversity, message=FALSE}
library(tidyverse)
tibble(Measure = alpha_diversities) %>%
  group_by(Measure) %>%
  summarize(Site1 = get(Measure)(s1), Site2 = get(Measure)(s2)) %>%
  pivot_longer(-Measure, "SampleID") %>%
  ggplot(aes(x=Measure, y=value, color=SampleID)) +
  geom_point() +
  scale_color_manual(values=c("#E64B35", "#4DBBD5")) +
  coord_flip() +
  theme_bw()
```

We can see that site 1 is regarded as more diverse by some measures; it has the
most species. For other measures, site 2 is regarded as more diverse; it has
the most even distribution of species.

In our documentation, you can find more info on each α-diversity
function.

Having assessed the within-sample diversity, we can next ask about the number
of shared species between sites.  If species are shared, how similar is the
distribution across species? There are many ways to quantify this between-sample
diversity or β-diversity.

We can view β-diversity as either the similarity or dissimilarity
between sites. The functions in `abdiv` are written in terms of
dissimilarity: similar sites will have values close to zero, and highly
dissimilar sites will have values close to the maximum.

The Jaccard distance counts the fraction of species present in only one site.
The answer is 3 out of 5, or 0.6.

```{r}
jaccard(s1, s2)
```

The Bray-Curtis dissimilarity adds up the absolute differences between species
counts, then divides by the total counts. For our two sites, that's
(2 + 5 + 8 + 8 + 7) / 48, or 0.625.

```{r}
bray_curtis(s1, s2)
```

Again, we'll use a vector called `beta_diversities` to compute every dissimilarity
measure in the library.

```{r beta-diversity}
tibble(Measure = beta_diversities) %>%
  group_by(Measure) %>%
  mutate(value = get(Measure)(s1, s2)) %>%
  ggplot(aes(x=Measure, y=value)) +
  geom_point(color="#4DBBD5") +
  scale_y_log10() +
  coord_flip() +
  theme_bw()
```

The dissimilarities are generally positive, and they have a range of scales.
Some dissimilarity measures range from 0 to 1, while others can go up
indefinitely.

As before, you can find more info on each β-diversity function in our documentation.
